Automatic Detection of Bad Programming Habits in Scratch: A Preliminary Study

### 2014-04-23 paperFIE_en.tex
Reviewing the paper. Changes here and there

%\documentclass[a4paper,10pt]{article}
\documentclass[conference]{format/IEEEtran}
%\usepackage[utf8x]{inputenc}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}

@@

\author{\IEEEauthorblockN{Jesús Moreno}
\IEEEauthorblockA{INTEF \& Programamos.es\\
Madrid, Spain\\
jesus.moreno@programamos.es}
\and
\IEEEauthorblockN{Gregorio Robles}
\IEEEauthorblockA{GSyC/LibreSoft\\Universidad Rey Juan Carlos\\
Madrid, Spain\\
grex@gsyc.urjc.es}}

@@

\begin{abstract}
%This paper describes a method for automatic detection of two bad programming habits that students learning to program with Scratch incur, such as the repetition of code and object naming, and raises ideas to try to avoid such situations. FIXME: some methodology, results and conclusions should be included in the abstract.
Using Scratch as a tool to teach Computer Science or develop Computational Thinking from other disciplines is increasingly common in all levels of education, and it is possible to find well-documented case studies from primary school to university.

However, there are reports indicating that students learning to program in this environment show certain habits that are contrary to the basic programming recommendations. In our work as instructors with high school students learning to program with Scratch, we have detected other bad practices, such as the repetition of code and object naming, that are repeated on a regular basis, and this study focuses on checking whether these issues are also present at the projects shared on the community website.

To test this hypothesys, we downloaded 100 projects from the Scratch website and analyzed them with two plug-ins we developed to automatically detect these bad practices. The plug-ins extend the functionality of the tool Hairball, a static code analyzer of Scratch projects.

The obtained results indicate that, in general, projects in the repository also incur in the investigated malpractices. Along with suggestions for future work, some ideas that might assist to avoid such situations are proposed in the conclusions of the paper.

\end{abstract}

@@

Scratch~\cite{resnick2009scratch}  is a programming environment including a visual programming language designed for children over 6 years old, a development environment and a website where the community can host their projects, run, study and reuse other programs, and share ideas or suggestions with other programmers. Although there are several similar visual programming languages, Scratch is undoubtedly the most successful one with more than three million registered users on its website and more than five million shared projects in its repository\footnote{See http://scratch.mit.edu/statistics/}.

Scratch is being used both in extracurricular activities~\cite{kafai2010entering} or summer camps~\cite{adams2010scratching}, as in elementary schools~\cite{wilson2012evaluation}, high schools~\cite{meerbaum2013learning} and even universities worldwide~\cite{malan2007scratch}. Its capacity to bring computer programming to children and teenagers with the aim of fostering computational thinking abilities and to promote ICT studies, and its success to teach both basic and advanced programming concepts (concepts such as user interaction, conditional statements, communication and synchronization, logical expressions and data representation are taught) has been documented~\cite{maloney2008programming, franklin2013assessment, wolz2009starting}.

@@

%FIXME: the goal of this paper is to... 
The goal of this paper, therefore, is to determine whether these bad habits are common practices in the Scratch community and propose some ideas to try to avoid such situations. In order to assist evaluators to detect these malpractices we developed two plug-ins that automate their detection by extending the functionality of Hairball, a static code analyzer of Scratch projects.

@@

Scratch also offers a website where the projects programmed by the community are published. The code of programs hosted there can be studied and analyzed, and it is possible to make suggestions and reuse the code, so there are lots of valuable resources available for anyone willing to start to program. This approach also offers the opportunity to learn certain social skills related to programming, such as sharing and contributing to the community~\cite{scaffidi2012skill}.

@@

Aiming to simplify the process of evaluating computational thinking and help evaluators with an automatic tool, Hairball was created~\cite{boe2013hairball}. Hairball is a static analyzer of Scratch projects inspired by Lint\footnote{http://www.unix.com/man-page/FreeBSD/1/lint} that tries to detect programming errors in Scratch projects, such as not initializing the value of a variable or the state of a character, sending messages that are not received by any program, including code that is never run, among others.

@@

In our work as instructors with high school students, we found that, despite our constant insistence and warnings to try to avoid them, many students show two bad habits on a regular basis that have to do with object naming and code repetition.

On the one hand, it is very common that students do not change the names of the characters, which are automatically named by the environment as \textit{SpriteX}, where X is an incremental number. When the amount of characters in a project increases, it is quite hard to know which object relates to a given statement, contributing to poor code readability, slowing the project programming and complicating debugging to fix bugs.

On the other hand, it is also very common to see students repeating code in the same project, sometimes even in the same character programs. Thus, besides complicating the maintenance and updating of the code, abstraction and modularization, two key components of the development of computational thinking~\cite{wing2008computational}, are not trained.

In order to detect these two behaviors in Scratch projects, we have developed two plug-ins for the Hairball environment~\cite{boe2013hairball}, which in turn makes use of \textit{kurt}\footnote{https://github.com/blob8108/kurt}, a library implemented in python, to automate the detection of this type of bad practices in programs developed in this environment.

The plug-in \textit{convention.SpriteNaming}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/convention.py} analyzes a Scratch project to check if the names of its characters begin with the string \textit{Sprite}, which indicates that the user has not changed its name and has left the one the system assigns by default to every new character added to the project.

The plug-in \textit{duplicate.DuplicateScripts}\footnote{https://github.com/jemole/hairball/blob/master/hairball/plugins/duplicate.py} analyzes a Scratch project to try to locate repeated complete programs within a project. In order to do so the plug-in runs each project and gets the \textit{tokens} of the different blocks of the programs, so that two blocks that only differ in the values they receive are considered to be equal. To illustrate the operation of this plug-in, picture \ref{fig:CodeRepetition1} shows two scripts containing the same blocks with the only difference of the values received by parameter.
\begin{figure}
  \centering
    \includegraphics[width=9cm]{img/CodeRepetition1.png}
  \caption{Two scripts repeating code}
  \label{fig:CodeRepetition1}
\end{figure}

When these scripts are analyzed by the plug-in and are translated into \textit{tokens} they would be considered equal, as their code would be translated to the next set of \textit{tokens}:
\begin{verbatim}
 'set rotation style %s'
 'repeat %s%s'
 'move %s steps'
 'next costume'
 'if on edge, bounce'
 'wait %s secs'
\end{verbatim}

Thus, the correct way to implement this project would have been as shown in Figure \ref{fig:CodeRepetition2}, defining a new set of blocks that receives several values per parameter and reusing the block created in both programs. This solution would involve improvements regarding maintenance, updating and debugging.

\begin{figure}
  \centering
    \includegraphics[width=9cm]{img/CodeRepetition2.png}
  \caption{Defining own blocks to avoid code repetition}
  \label{fig:CodeRepetition2}
\end{figure}

In order to test the functionality of the developed plug-ins and check whether these behaviors that we detected in our work are also common in the projects of the Scratch repository, we proceeded to download 100 projects from the repository. As the possibility for users to define their own blocks, using \textit{def\_block} feature, is only available since the release of Scratch 2.0, we selected and downloaded projects created after that date.

@@

Data obtained after analyzing 100 downloaded projects confirm that observations made with our students are also very common in the community projects.

Regarding the lack of habit of renaming the characters and not using the default name assigned by the environment, the 79\% of the analyzed projects contained at least one \textit{SpriteX} character not renamed.

As an example to illustrate the drawbacks associated with this malpractice, picture \ref{fig:SpriteNaming} shows a script that is difficult to read as the character area must be explored to find which character refers to a particular block, making the program difficult to debug any problems should arise, making it hard to code improvements and even slowing down the programming time, as the name of the character must be memorized or checked every time a block is added.

@@

@@

This paper shows the preliminary results of a study regarding two bad programming habits we have detected in our work as instructors with high school students learning to program with Scratch. In order to check if these bad habits, which have to do with object naming and code repetition, are also commonly found in the projects shared in the community repository, we downloaded 100 proyects and analyzed them with two plug-ins we developed for Hairball, detecting that most of the inspected proyects, 79\% and 62\%  respectively, fall into these issues.

Regarding the object naming, it is curious that Scratch users name correctly, i.e., semantically meaningful, the variables used in their projects. In our opinion the reason to explain this paradox is because it is mandatory to assign a name to a variable the moment it is created, as the environment does not name it by default. However, when creating a new character a  \textit{SpriteX} name is assigned automatically. In our views, a change of this feature of the environment, forcing users to name new objects, might eliminate this bad practice.

Regarding the code repetition and the low usage Scratch users make of the possibility of creating their own blocks (the equivalent of defining methods or procedures in other languages), works as the one performed by Seiter and Foreman~\cite {seiter2013modeling}, in which they show that abstraction and modularization capacities are concepts that seem to be developed from a certain age, may explain the low rates of use of this functionality, since a significant part of the Scratch community consists of children under 10 years old\footnote{http://scratch.mit.edu/statistics/\#age}. Future studies should further investigate to continue advancing in the definition of an appropriate framework for different ages, as well as to check whether users learning online and those who do it in a regulated environment show different use rates of this functionality.

@@

In the near future we plan to extend the scope of our study developing new plug-ins to automate the detection of other issues, such as the degree of development of computational thinking demonstrated in a project, and analyzing a much larger number of projects to achieve more accurate conclusions. In this sense, the release of a research dataset with the first five years of data from the Scratch repository\footnote{https://sites.google.com/site/scratchdatameeting/} will be certainly a great help to our work. Finally, we are in the process of developing a website that will allow students to receive feedback from each of its programs with warnings of potential errors and tips to improve their code, and will offer teachers a tool to assist them in correcting their students proyects.

@@

The work of Gregorio Robles has
been funded in part by the Spanish Government under project 
SobreSale (TIN2011-28110) and by the Comunidad de Madrid 
under project eMadrid (FIXME).

>>>>

%\documentclass[a4paper,10pt]{article}
\documentclass[conference]{format/IEEEtran}
%\usepackage[utf8x]{inputenc}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{url}

@@

\author{\IEEEauthorblockN{Jesús Moreno}
\IEEEauthorblockA{Programamos.es \& Instituto Nacional de \\ Tecnologías Educativas y de Formación del Profesorado\\
Madrid, Spain\\
jesus.moreno@programamos.es}
\and
\IEEEauthorblockN{Gregorio Robles}
\IEEEauthorblockA{GSyC/LibreSoft\\Universidad Rey Juan Carlos\\
Madrid, Spain\\
grex@gsyc.urjc.es}}

@@

\begin{abstract}
%This paper describes a method for automatic detection of two bad programming habits that students learning to program with Scratch incur, such as the repetition of code and object naming, and raises ideas to try to avoid such situations. FIXME: some methodology, results and conclusions should be included in the abstract.
Using the Scratch environment as a tool to teach programming skills or develop computational thinking is increasingly common in all levels of education, well-documented case studies from primary school to university can be found.

However, there are reports that indicate that students learning to program in this environment show certain habits that are contrary to the basic programming recommendations. In our work as instructors with high school students, we have detected these and other bad practices, such as the repetition of code and object naming, on a regular basis. This paper focuses on verifying whether these issues can be generally be found in the Scratch community, by analyzing a large number of projects available on the Scractch community website.

To test this hypothesis, we downloaded 100 projects and analyzed them with two plug-ins we developed to automatically detect these bad practices. The plug-ins extend the functionality of the Hairball tool, a static code analyzer for Scratch projects.

The results obtained show that, in general, projects in the repository also incur in the investigated malpractices. Along with suggestions for future work, some ideas that might assist to address such situations are proposed in the conclusions of the paper.
\end{abstract}

@@

Scratch~\cite{resnick2009scratch}  is a programming environment including a visual programming language designed for children over 6 years old, a development environment and a website where the community can host their projects, run, study and reuse other programs, and share ideas or suggestions with other programmers. Although there are several similar visual programming languages, Scratch is undoubtedly the most successful one with more than three million registered users in its website and more than five million shared projects in its repository\footnote{See \url{http://scratch.mit.edu/statistics/}}.

Scratch is being used both in extracurricular activities~\cite{kafai2010entering} or summer camps~\cite{adams2010scratching}, as in elementary schools~\cite{wilson2012evaluation}, high schools~\cite{meerbaum2013learning} and even universities worldwide~\cite{malan2007scratch}. Its capacity to bring computer programming to children and teenagers with the aim of fostering computational thinking abilities and to promote ICT studies, and its success to teach both basic and advanced programming concepts (concepts such as user interaction, conditional statements, communication and synchronization, logical expressions and data representation are taught) has been well documented~\cite{maloney2008programming, franklin2013assessment, wolz2009starting}.

@@

%FIXME: the goal of this paper is to... 
The goal of this paper, therefore, is to determine whether these bad habits are common practices in the Scratch community in an automatic manner and propose some ideas to try to avoid such situations. In order to assist evaluators to detect these malpractices we have developed two plug-ins that automate the detection of the bad programming practices by extending the functionality of Hairball, a static code analyzer of Scratch projects.

@@

Scratch also offers a website where the projects programmed by the community are published. The code of programs hosted there can be run, and following the community idea of the free software movement it can be studied and analyzed. So, it is possible to make suggestions and to reuse the code, and there are lots of valuable resources available for anyone willing to start to program. This approach offers as well the opportunity to learn social skills related to programming and beyond, such as sharing and contributing to the community~\cite{scaffidi2012skill}.

@@

Aiming to simplify the process of evaluating computational thinking and help evaluators with an automatic tool, Hairball was created~\cite{boe2013hairball}. Hairball is a static analyzer of Scratch projects inspired by Lint\footnote{\url{http://www.unix.com/man-page/FreeBSD/1/lint}} that tries to detect programming errors in Scratch projects, such as not initializing the value of a variable or the state of a character, sending messages that are not received by any program, including code that is never run, among others.

@@

In our work as instructors with high school students, we found that, despite our constant insistence and warnings, many students show on a regular basis two bad habits that have to do with object naming and code repetition:

\begin{enumerate}

  \item Students do not change the names of the characters, which are automatically named by the environment as \textit{SpriteX} where X is an incremental number. When the amount of characters in a project is large, it becomes very hard to know what object relates to a given statement, contributing to poor code readability, slowing the project programming, and complicating debugging to fix bugs.

  \item Students repeat code in the same project, sometimes even in the same character programs. Thus, abstraction and modularization, two key components of the development of computational thinking~\cite{wing2008computational}, are not trained. And the \emph{code base} becomes more complicated to maintain and update in the future.

\end{enumerate}

In order to detect these two behaviors in Scratch projects automatically, we have developed following plug-ins for the Hairball environment~\cite{boe2013hairball}, which in turn make use of \textit{kurt}\footnote{\url{https://github.com/blob8108/kurt}}, a library implemented in Python. The developed plug-ins are:

\begin{enumerate}

  \item \textit{convention.SpriteNaming}\footnote{\url{https://github.com/jemole/hairball/blob/master/hairball/plugins/convention.py}} analyzes a Scratch project to check if the names of its characters begin with the string \textit{Sprite}, indicative for the user not having changed the default name.

As an example to illustrate the drawbacks associated with this bad programming behavior, Fig.~\ref{fig:SpriteNaming} shows a script that is difficult to read because the complete character area has to be explored to find what character refers to a particular block. This makes the program difficult to debug should errors arise, making it hard to improve the code and slowing down the programming tasks.

  \item \textit{duplicate.DuplicateScripts}\footnote{\url{https://github.com/jemole/hairball/blob/master/hairball/plugins/duplicate.py}} analyzes a Scratch project to try to locate repeated complete programs within a project. In order to do so, the plug-in runs each project and gets the \textit{tokens} of the different blocks of the programs, so that two blocks that only differ in the values they receive are considered to be equal. To illustrate the operation of this plug-in, Fig.~\ref{fig:CodeRepetition1} shows two scripts containing the same blocks with the only difference of the values received by parameter.

\begin{figure}
  \centering
    \includegraphics[width=9cm]{img/CodeRepetition1.png}
  \caption{Two scripts repeating code}
  \label{fig:CodeRepetition1}
\end{figure}

The Scratch scripts are analyzed by the plug-in and translated into \textit{tokens}. The analysis would result in considering them equal, as their code would be translated in both cases into the following sequence of \textit{tokens}:
\begin{verbatim}
 'set rotation style %s'
 'repeat %s%s'
 'move %s steps'
 'next costume'
 'if on edge, bounce'
 'wait %s secs'
\end{verbatim}

Hence, the correct way to implement this functionality would have been as shown in Fig.~\ref{fig:CodeRepetition2}, defining a new set of blocks that receives several values per parameter, and then reusing the block created in both programs. This solution would facilitate future maintenance, updating and debugging.

\begin{figure}
  \centering
    \includegraphics[width=9cm]{img/CodeRepetition2.png}
  \caption{Definition of blocks to avoid code repetition.}
  \label{fig:CodeRepetition2}
\end{figure}

\end{enumerate}

In order to test the functionality of the developed plug-ins and to verify whether these behaviors are common in projects found in the Scratch repository, we proceeded to randomly download 100 projects from the repository. As the possibility for users to define their own blocks, using \textit{def\_block} feature, is only available since the release of Scratch 2.0, we selected and downloaded projects created for this version. A replication package of the projects under study, and the specific version of the plug-ins used in it can be obtained publicly\footnote{FIXME: URL \url{http://gsyc.urjc.es/~grex/repro/2014-fie-scratch}}.

@@

The results obtained after analyzing the 100 projects confirm that observations made with our students can also be generally found in the community projects.

Regarding the lack of habit of renaming the characters and not using the default name assigned by the environment, 79\% of the analyzed projects contained at least one non-renamed \textit{SpriteX} character.

@@

FIXME: here comes the table with some more results. Important: to note how the distribution (talk about mean and median).

@@

This paper shows the preliminary results of a study regarding two bad programming habits we have detected in our work as instructors with high school students learning to program with Scratch. In order to check if these bad habits, which have to do with object naming and code repetition, are also commonly found in the projects shared in the community repository, we downloaded 100 projects and analyzed them with two plug-ins we developed for Hairball, detecting that most of the inspected projects, 79\% and 62\%  respectively, fall into these issues.

Regarding object naming, it is worth to note that Scratch users name the variables used in their projects correctly, i.e., semantically meaningful. In our opinion the reason to explain this paradox is because it is mandatory to assign a name to a variable the moment it is created, as the environment does not name it by default. However, when creating a new character a  \textit{SpriteX} name is assigned automatically. In our views, a change of this feature of the environment, forcing users to name new objects, might eliminate this bad practice.

Regarding code repetition and the low usage Scratch users make of the possibility of creating their own blocks (the equivalent of defining methods or procedures in other languages), works as the one performed by Seiter and Foreman~\cite {seiter2013modeling}, in which they show that abstraction and modularization capacities are concepts that seem to be developed from a certain age, may explain the low rates of use of this functionality, since a significant part of the Scratch community consists of children under 10 years\footnote{\url{http://scratch.mit.edu/statistics/#age}}. Future studies should further investigate how to continue advancing in the definition of an appropriate framework for different ages, as well as to check whether users learning online and those who do it in a regulated environment show different use rates of this functionality.

@@

In the near future we plan to extend the scope of our study developing new plug-ins to automate the detection of other issues, such as the degree of development of computational thinking demonstrated in a project, and by analyzing a much larger number of projects to achieve more accurate conclusions. In this sense, the release of a research dataset with the first five years of data from the Scratch repository\footnote{\url{https://sites.google.com/site/scratchdatameeting/}} will be certainly a great help to our work. Finally, we are in the process of developing a website that will allow students to receive immediate feedback for their programs with warnings of potential errors and tips to improve their code.

@@

The work of Gregorio Robles has
been funded in part by the Spanish Government under project 
SobreSale (TIN2011-28110) and under project ``eMadrid - Investigación y Desarrollo de tecnologías para el e-learning en la Comunidad de Madrid'' (S2009/TIC-1650) funded by the Region of Madrid.
